Environmental pathogen reservoirs and habitat heterogeneity in a metapopulation

### 2014-04-23 Manuscript/manuscript.tex
Introduction edits
Attempts to streamline introduction.

Patterns and rates of host movement also play an important role in pathogen spread through a metapopulation.  As shown by \cite{Hess1996}, for pathogens spread by direct contact alone, increased host movment enhances pathogen spread through a metapopulation.  Further studies by \cite{Gog2002} and \cite{McCallum2002} explore the role of additional transmission pathways by including background infection from an alternate host, while \cite{Park2012} includes transmission from an environmental reservoir.  These studies show that the the effect of host movement on occupancy (and pathogen spread) depends on the relative strength of the different transmission pathways and highlight the potential for patterns of host movement and pathways of pathogen transmission to interact and influence disease dynamics.

However, the above studies assume that all patches in the metapopulation are identical, and consider movement only at the metapopulation scale.  Since environmental transmission is spatially explicit (i.e. environmental reservoirs can only infect residents of the local patch), its role in disease dynamics may depend on the attributes of the specific habitat patches in which reservoirs become established.  In particular, the quality of a habitat patch can affect its extinction and colonization probabilities, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).  As an example, in a study of the Glanville fritillary butterfly metapopulation, \cite{Moilanen1998} found that increased patch quality, as measured by the density of flowering plants, decreased emigration and increased immigration.  Thus, patterns of host movement, and consequently the role of environmental pathogen transmission, are determined at least in part by patch quality.

Many of the pathogens listed above affect spatially structured host populations (e.g. plague in prairie dog colonies), and thus understanding the consquences of environmental pathogen persistence and habitat quality for pathogen spread is critical to managing disease in these systems.  In this study, we ask how pathogen longevity (how long the pathogen can persist and remain infections in the environment), and habitat heterogeneity (the variance in the distribution of patch quality across the metapopulation) interact to influence pathogen spread through the metapopulation.  Within a metapopulation, we ask what specific role low and high quality patches play in controlling pathogen spread.

We expect pathogen spread to increase with both environmental longevity and with habitat heterogeneity.  Increased environmental longevity should provide greater opportunities for transmission, and thus should favor pathogen spread.  The role of habitat heterogeneity is likely to be more complex, emerging from the behavior of high and low quality patchs in heterogeneous metapopulations.  Specifically, we expect that high quality habitat patches, which support greater host traffic than lower quality habitat, might be more likely to form pathogen reservoirs.  As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (citations, examples?).  In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (citation).  Similarly, we predict that low quality patches, which see relatively less host traffic, will be less likely to develop pathogen reservoirs and and thus may serve as refuges for susceptible hosts to escape infection.

>>>>

Since environmental transmission is spatially explicit (i.e. environmental reservoirs can only infect residents of the local patch), its role in disease dynamics may depend on the attributes of the specific habitat patches in which reservoirs become established.  In particular, the quality of a habitat patch can affect its extinction and colonization probabilities, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).  Thus, we expect that patterns of host movement, and consequently the effect of environmental pathogen transmission, are determined at least in part by patch quality.

Many of the pathogens listed above affect spatially structured host populations (e.g. plague in prairie dog colonies), and thus understanding how environmental pathogen persistence interacts with habitat quality is critical to managing disease in these systems.  In this study, we ask how pathogen longevity (how long the pathogen can persist and remain infections in the environment), and habitat heterogeneity (the variance in the distribution of patch quality across the metapopulation) interact to influence pathogen spread and metapopulation occupancy.  Within a metapopulation, we ask what specific role low and high quality patches play in controlling pathogen spread.

We expect pathogen spread to increase with both environmental longevity and with habitat heterogeneity.  Increased environmental longevity should provide greater opportunities for transmission, and thus should favor pathogen spread.  The role of habitat heterogeneity is likely to be more complex, emerging from the behavior of high and low quality patchs in heterogeneous metapopulations.  Specifically, we expect that high quality habitat patches, which support greater host traffic than lower quality habitat, might be more likely to form pathogen reservoirs.  As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Almberg2011}, citations, examples?).  In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (citation).  Similarly, we predict that low quality patches, which see relatively less host traffic, will be less likely to develop pathogen reservoirs and and thus may serve as refuges for susceptible hosts to escape infection.
